%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%								%
%								%
%								%
%			  biblio.tex				% 
%								%
%								%
%								%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{00}

\bibitem{AB} A.~Aldini and M.~Bernardo,
{\it ``On the Usability of Process Algebra: An Architectural View''},
in Theoretical Computer Science~335:281-329, 2005.

\bibitem{BBS} S.~Balsamo, M.~Bernardo, and M.~Simeoni,
{\it ``Performance Evaluation at the Software Architecture Level''},
in {\it Formal Methods for Software Architectures},
LNCS~2804:207-258, 2003.

\bibitem{BSW} K.A.~Bartlett, R.A.~Scantlebury, and P.T.~Wilkinson,
{\it ``A Note on Reliable Full-Duplex Transmission over Half-Duplex Links''},
in Comm.\ of the ACM~12:260-261, 1969.

\bibitem{Ber1} M.~Bernardo,
{\it ``Theory and Application of Extended Markovian Process Algebra''},
Ph.D.\ Thesis, University of Bologna (Italy), 1999.

\bibitem{Ber2} M.~Bernardo,
{\it ``Symbolic Semantic Rules for Producing Compact STGLA from Value Passing Process Descriptions''},
in ACM Trans.\ on Computational Logic~5:436-469, 2004.

\bibitem{Ber3} M.~Bernardo,
{\it ``Weak Markovian Bisimilarity for GSPNs and EMPA$_{\rm gr}$''},
submitted for publication.

\bibitem{BB1} M.~Bernardo and M.~Bravetti,
{\it ``Performance Measure Sensitive Congruences for Markovian Process Algebras''},
in Theoretical Computer Science~290:117-160, 2003.

\bibitem{BCD} M.~Bernardo, P.~Ciancarini, and L.~Donatiello,
{\it ``Architecting Families of Software Systems with Process Algebras''},
in ACM Trans.\ on Software Engineering and Methodology~11:386-426, 2002.

\bibitem{BDC} M.~Bernardo, L.~Donatiello, and P.~Ciancarini,
{\it ``Stochastic Process Algebra: From an Algebraic Formalism to an Architectural Description Language''},
in {\it Performance Evaluation of Complex Systems: Techniques and Tools},
LNCS~2459:236-260, 2002.

\bibitem{BB2} M.~Bravetti and M.~Bernardo,
{\it ``Compositional Asymmetric Cooperations for Process Algebras with Probabilities, Priorities, and
Time''},
in Proc.\ of the {\it 1st Int.\ Workshop on Models for Time Critical Systems (MTCS~2000)},
ENTCS~39(3),
State College (PA), 2000.

\bibitem{Buc} P.~Buchholz,
{\it ``Exact and Ordinary Lumpability in Finite Markov Chains''},
in Journal of Applied Probability~31:59-75, 1994.

\bibitem{CCOPR} R.~Cavada, A.~Cimatti, E.~Olivetti, M.~Pistore, and M.~Roveri,
{\it ``NuSMV~2.1 User Manual''},
2002.

\bibitem{CGH} G.~Clark, S.~Gilmore, and J.~Hillston,
{\it ``Specifying Performance Measures for PEPA''},
in Proc.\ of the {\it 5th AMAST Int.\ Workshop on Formal Methods for Real Time and Probabilistic Systems
(ARTS~1999)},
LNCS~1601:211-227, 1999.

\bibitem{CGP} E.M.~Clarke, O.~Grumberg, and D.A.~Peled,
{\it ``Model Checking''},
MIT Press, 1999.

\bibitem{Cle} W.R.~Cleaveland,
{\it ``On Automatically Explaining Bisimulation Inequivalence''},
in Proc.\ of the {\it 2nd Int.\ Conf.\ on Computer Aided Verification (CAV~1990)},
LNCS~531:364-372,
New Brunswick (NJ), 1990.

\bibitem{CLS} W.R.~Cleaveland, T.~Li, and S.~Sims,
{\it ``The Concurrency Workbench of the New Century -- Version~1.2''},
2000.

\bibitem{CS} W.R.~Cleaveland and O.~Sokolsky,
{\it ``Equivalence and Preorder Checking for Finite-State Systems''},
in {\it Handbook of Process Algebra},
Elsevier, pp.~391-424, 2001.

\bibitem{DH} R.~De~Nicola and M.C.B.~Hennessy,
{\it ``Testing Equivalences for Processes''},
in Theoretical Computer Science~34:83-133, 1983.

\bibitem{FG} R.~Focardi and R.~Gorrieri,
{\it ``A Classification of Security Properties''},
in Journal of Computer Security~3:5-33, 1995.

\bibitem{HM} M.C.B.~Hennessy and R.~Milner,
{\it ``Algebraic Laws for Nondeterminism and Concurrency''},
in Journal of the ACM~32:137-161, 1985.

\bibitem{Hil} J.~Hillston,
{\it ``A Compositional Approach to Performance Modelling''},
Cambridge University Press, 1996.

\bibitem{How} R.A.~Howard,
{\it ``Dynamic Probabilistic Systems''},
John Wiley \& Sons, 1971.

\bibitem{Jai} R.~Jain,
{\it ``The Art of Computer Systems Performance Analysis''},
John Wiley \& Sons, 1991.

\bibitem{KS} P.C.~Kanellakis and S.A.~Smolka,
{\it ``CCS Expressions, Finite State Processes, and Three Problems of Equivalence''},
in Information and Computation~86:43-68, 1990.

\bibitem{KMM} M.H.~Kang, A.P.~Moore, and I.S.~Moskowitz,
{\it ``Design and Assurance Strategy for the NRL Pump''},
in IEEE Computer Magazine~31:56-64, 1998.

\bibitem{KR} B.W.~Kernighan and D.M.~Ritchie,
{\it ``The C Programming Language''},
Prentice Hall, 1988.

\bibitem{LS} K.G.~Larsen and A.~Skou,
{\it ``Bisimulation through Probabilistic Testing''},
in Information and Computation~94:1-28, 1991.

\bibitem{LR} D.~Lehmann and M.~Rabin,
{\it ``On the Advantage of Free Choice: A Symmetric and Fully Distributed Solution to the Dining
Philosophers Problem''},
in Proc.\ of the {\it 8th Symp.\ on Principles of Programming Languages (POPL~1981)},
ACM Press, pp.~133-138,
New York (NY), 1981.

\bibitem{Mil} R.~Milner,
{\it ``Communication and Concurrency''},
Prentice Hall, 1989.

\bibitem{Ous} J.K.~Ousterhout,
{\it ``Tcl and the Tk Toolkit''},
Addison-Wesley, 1994.

\bibitem{SG} M.~Shaw and D.~Garlan,
{\it ``Software Architecture: Perspectives on an Emerging Discipline''},
Prentice Hall, 1996.

\bibitem{Ste1} W.J.~Stewart,
{\it ``Introduction to the Numerical Solution of Markov Chains''},
Princeton University Press, 1994.

\bibitem{Ste2} W.J.~Stewart,
{\it ``MarCA: Markov Chain Analyzer -- Version~3.0''},
1996.

\bibitem{Wel} P.D.~Welch,
{\it ``The Statistical Analysis of Simulation Results''},
in {\it ``Computer Performance Modeling Handbook''},
Academic Press, pp.~267-329, 1983.

\end{thebibliography}
